# Adverse Impact of Severe Weather Events on Public Health and Economy Between 1950 and 2011

## Synopsis
Severe weather events have a potential to cause widespread adverse impact to both public health as well as economy.  The objective of this analysis is to explore the U.S. National Oceanic and Atmospheric Administrations's (NOAA) storm database, that has data from year 1950 to November 2011, in order to find out which severe weather events caused most harm to health both in terms of injury and fatalities and which severe weather events cause the most damage in economic terms.  To answer these questions we process data from NOAA database to find top fifteen weather events that caused most harm to health as well as most economic damage.  The analysis conducted here indicates that while Tornado caused most harm to health, both in terms of injuries and fatalities individually as well as combined, Flood was the most damaging weather event for economy. 

## Data Processing
```{r globalOptions}
## Set global options
opts_chunk$set(echo=TRUE, cache=FALSE)
```

Load, explore and subset data
```{r load}
## download file
filesrc <- 
"http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
filename <- "repdata-data-StormData.csv.bz2"
download.file(filesrc, destfile = filename)
## Reading the file
data <- read.csv(bzfile(filename), sep=",",quote="\"", na.strings = "NA")
## Getting sense of data
dim(data)
head(data)
str(data)
colnames(data)
## Subset data
subData <- data[, c("STATE", "EVTYPE", "FATALITIES", "INJURIES", 
               "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
head(subData)
```

Remove inconsequential values
```{r expldata}
## Check to see if there are rows that have value for all four columns
## FATALITIES and INJURIES and PROPDMG and CROPDMG is zero
ignore <- subData[which(subData$FATALITIES == 0 & subData$INJURIES == 0 &
                         subData$PROPDMG == 0 & subData$CROPDMG == 0), ]
## Ignore these rows that have values for all four columns to be zero because
## for the purpose of this analysis, these rows are inconsequential for their
## corresponding event types as they did not lead to any damage health or
## property.  So, select all rows that have non-zero value for at least one of
## these four columns.
myData <- subData[which(subData$FATALITIES > 0 | subData$INJURIES > 0 |
                         subData$PROPDMG > 0 | subData$CROPDMG > 0), ]
```

Clean event types according to specification in document.

Here we are looking to clean events that can be categrorized as TORNADO.
```{r checkcleantornado}
tornado <- myData[grepl("torn", myData$EVTYPE, ignore.case = TRUE),]
#Check to see if there are any event types to be cleaned
chktornado <- myData[grepl("torn", myData$EVTYPE, ignore.case = TRUE) & 
                             !grepl("tornado", 
                                    myData$EVTYPE, ignore.case = TRUE),]
## There is one event type where TORNADO  is misspelled as TORNDAO.  Clean in
## by substituting with correct spelling
myData[grepl("TORNDAO", 
             myData$EVTYPE, ignore.case = TRUE), "EVTYPE"] <- "TORNADO"
myData[grepl("TORNTORNDAO", 
             myData$EVTYPE, ignore.case = TRUE), "EVTYPE"] <- "TORNADO"
## Clean observations as:
## TORNADOES, TSTM WIND, HAIL -> TORNADO,
## TORNADOES -> TORNADO
myData[grepl("tornadoes", 
             myData$EVTYPE, ignore.case = TRUE), "EVTYPE"] <- "TORNADO"
myData[grepl("COLD AIR TORNADO", 
             myData$EVTYPE, ignore.case = TRUE), "EVTYPE"] <- "TORNADO"
myData[grepl("TORNADO F", 
             myData$EVTYPE, ignore.case = TRUE), "EVTYPE"] <- "TORNADO"
```

Categorize Avalance related events.
```{r checkcleanavlnch}
avl <- myData[grepl("aval", myData$EVTYPE, ignore.case = TRUE) &
                      !grepl("Avalanche", myData$EVTYPE, ignore.case = TRUE),]
## One entry where Avalanche is mispelled as Avalance.  Clean it up by 
## replacing with correct spelling.
myData$EVTYPE <- sub("AVALANCE", "AVALANCHE", myData$EVTYPE)
```

Categorize BLIZZARD related events.
```{r checkcleanblzrd, cache=TRUE}
## Explore blizzard related data
bl <- myData[grepl("bli", myData$EVTYPE, ignore.case = TRUE), ]
blz <- myData[grepl("bli", myData$EVTYPE, ignore.case = TRUE) &
                      !grepl("Blizzard", myData$EVTYPE, ignore.case = TRUE),]
myData[grepl("BLIZZARD/WINTER STORM", myData$EVTYPE, 
             ignore.case = TRUE), "EVTYPE"] <- "BLIZZARD"
myData[grepl("GROUND BLIZZARD", myData$EVTYPE, 
             ignore.case = TRUE), "EVTYPE"] <- "BLIZZARD"
## All rows for blizzard are already correct.  No cleaning needed.
```

Check and clean COASTAL FLOOD related events.
```{r chkclncoastalflood, cache=TRUE}
cf <- myData[grepl("^coastal flood", myData$EVTYPE, ignore.case = TRUE), ]
## All rows for Coastal flood are spelled correct.  But there are some rows
## that contain Coastal flooding and some rows have coastal flood/erosion instead
## of Coastal Flood that is specified in the in documentaion.  Clean it so that
## all these event types comply to the accompanying code book and so have value
## Coastal flood
myData$EVTYPE <- sub("coastal flooding", "Coastal Flood", myData$EVTYPE,
                     ignore.case = TRUE)
myData$EVTYPE <- sub("coastal flooding/erosion", "Coastal Flood", myData$EVTYPE,
                     ignore.case = TRUE)
myData$EVTYPE <- sub("coastal flood/erosion", "Coastal Flood", myData$EVTYPE,
                     ignore.case = TRUE)
```

Work on events related to COLD/WIND CHILL.
```{r chkclncoldwindchill}
cwc <- myData[grepl("^cold/", myData$EVTYPE, ignore.case = TRUE), ]
## There is one observation with COLD/WINDS that does not comply with EVTYPE
## value in code book.  The actual value is COLD/WIND CHILL.  Replace with 
## correct value.
myData$EVTYPE <- sub("^cold/winds", "COLD/WIND CHILL", myData$EVTYPE, 
                     ignore.case = TRUE)
uev <- unique(myData$EVTYPE)
ucld <- uev[grepl("cold", uev, ignore.case = TRUE)]
myData[grepl("cold", myData$EVTYPE, 
                    ignore.case = TRUE) &
               !grepl("COLD/WIND CHILL", myData$EVTYPE, 
                    ignore.case = TRUE) &
               !grepl("EXTREME COLD/WIND CHILL", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "COLD/WIND CHILL"
```
Check and clean events related to DEBRIS FLOW.
```{r chkclndebrisflow}
deb <- myData[grepl("^deb", myData$EVTYPE, ignore.case = TRUE), ]
## No rows for Debris flow
```
Work on events related to DENSE FOG.
```{r chkclndnsfg}
dfg <- myData[grepl("^dens", myData$EVTYPE, ignore.case = TRUE), ]
chkdfg <- myData[grepl("^dens", myData$EVTYPE, ignore.case = TRUE) &
                         !grepl("^dense fog", myData$EVTYPE, 
                                ignore.case = TRUE), ]
## So all the observations for EVTYPE DENSE FOG have values complying to 
## code book.  we also find out that there is only one entry for 
## DENSE SMOKE and that complies to code book too.  So no cleaning required
## for DENSE FOG or DENSE SMOKE
```
Work on evens related to DROUGHT.
```{r chkclndrought}
drt <- myData[grepl("drou", myData$EVTYPE, ignore.case = TRUE), ]
## There are some observations that have EVTYPE value of DROUGHT/EXCESSIVE HEAT.
## Both DROUGHT and EXCESSIVE HEAT are valid values for EVTYPE according to 
## code book and there is no clear cut direction as to how to deal with this
## situation.  For this analysis, I am considering it be DROUGHT.  That implies 
## we need to replace DROUGHT/EXCESSIVE HEAT with DROUGHT.
myData$EVTYPE <- sub("DROUGHT/EXCESSIVE HEAT", "DROUGHT", myData$EVTYPE, 
                     ignore.case = TRUE)
drtcln <- myData[grepl("drou", myData$EVTYPE, ignore.case = TRUE) &
                         !grepl("^drou", myData$EVTYPE, ignore.case = TRUE), ]
## There is one entry that has DROUGHT at the end.  Since there is no
## EVTYPE with value HEAT WAVE, so this needs to be cleaned so that it gets
## included in DROUGHT
myData$EVTYPE <- sub("HEAT WAVE DROUGHT", "DROUGHT", myData$EVTYPE, 
                     ignore.case = TRUE)
````
Work on events related to DUST DEVIL.
```{r chkclndstdvl}
dd <- myData[grepl("dust devil", myData$EVTYPE, ignore.case = TRUE), ]
## There is one observation that has value DUST DEVIL WATERSPOUT.  For this
## analysis this entry is being considered as DUST DEVIL.  So, clean this
## value
myData$EVTYPE <- sub("DUST DEVIL WATERSPOUT", "DUST DEVIL", 
                     myData$EVTYPE, ignore.case = TRUE)
ddchk <- myData[grepl("dust devil", myData$EVTYPE, ignore.case = TRUE) &
                     !grepl("^dust", myData$EVTYPE, ignore.case = TRUE) , ]
```
Work on events related to DUST STORM.
```{r chkclndststol}
## Check and clean DUST STORM
ds <- myData[grepl("dust storm", myData$EVTYPE, ignore.case = TRUE), ]
dschk <- myData[grepl("dust", myData$EVTYPE, ignore.case = TRUE) &
                        !grepl("^dust", myData$EVTYPE, ignore.case = TRUE), ]
## There is an observation that has value BLOWING DUST and another having value
## DUST STORM/HIGH WINDS.  Both of these need to be replaced by DUST STORM
myData$EVTYPE <- sub("BLOWING DUST", "DUST STORM", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("DUST STORM/HIGH WINDS", "DUST STORM", 
                     myData$EVTYPE, ignore.case = TRUE)
```
Work on events related to EXCESSIVE HEAT.
```{r chkclnexht}
eh <- myData[grepl("excessive heat", myData$EVTYPE, ignore.case = TRUE), ]
## There is an observation with EXTREME HEAT that does not belong here
ehchk <- myData[grepl("excessive heat", myData$EVTYPE, ignore.case = TRUE) &
                         !grepl("^excessive heat", myData$EVTYPE, 
                                ignore.case = TRUE), ]
## There is one observation with RECORD/EXCESSIVE HEAT that needs to be 
## made compliant with code book, so it should be replaced with EXCESSIVE HEAT
myData$EVTYPE <- sub("RECORD/EXCESSIVE HEAT", "EXCESSIVE HEAT", 
                     myData$EVTYPE, ignore.case = TRUE)
ehp <- myData[grepl("excessive heat", myData$EVTYPE, ignore.case = TRUE) &
                      !grepl("^excessive heat", myData$EVTYPE, 
                                ignore.case = TRUE), ]
```
Work on events related to EXTREME COLD/WIND CHILL.
```{r chkclnextcld}
ec <- myData[grepl("extr", myData$EVTYPE, ignore.case = TRUE), ]
## There is an observation with EXTREME HEAT that does not belong here
echks <- myData[grepl("extre", myData$EVTYPE, ignore.case = TRUE) &
                         !grepl("^extre", myData$EVTYPE, ignore.case = TRUE), ]
## There is an observation with EXTREME HEAT that does not belong here
echk <- myData[grepl("extre", myData$EVTYPE, ignore.case = TRUE) &
                         !grepl("extreme cold", myData$EVTYPE, 
                                ignore.case = TRUE), ]
## There are a few observations with value EXTREME HEAT, EXTREME WIND CHILL
## and EXTREME WINDCHILL.  Whereas, EXTREME HEAT does not belong here,
## the other two values need to be replaced with the correct value from code
## book, which is - EXTREME COLD/WIND CHILL
myData$EVTYPE <- sub("EXTREME WIND CHILL", "EXTREME COLD/WIND CHILL", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("EXTREME WINDCHILL", "EXTREME COLD/WIND CHILL", 
                     myData$EVTYPE, ignore.case = TRUE)
## Replace all values with EXTREME COLD to EXTREME COLD/WIND CHILL which is
## the correct value according to code book
myData[grepl("EXTREME COLD", myData$EVTYPE, ignore.case = TRUE),
       "EVTYPE"] <- "EXTREME COLD/WIND CHILL"
```
Work on events related FLASH FLOOD.
```{r chkclnff}
fl <- myData[grepl("flash", myData$EVTYPE, ignore.case = TRUE), ]
## There is an observation with ICE STORM/FLASH FLOOD.  Both of these are
## valid values according to the code book.  For this analysis treat this as 
## ICE STORM.  Also, there is value with FLASH FLOODING.  This needs to be 
## replaced with FLASH FLOOD.
myData$EVTYPE <- sub("FLASH FLOODING", "FLASH FLOOD", 
                     myData$EVTYPE, ignore.case = TRUE)
ffp <- myData[grepl("flash", myData$EVTYPE, ignore.case = TRUE), ]
ffchk <- myData[grepl("flash", myData$EVTYPE, ignore.case = TRUE) &
                         !grepl("^flash", myData$EVTYPE, 
                                ignore.case = TRUE), ]
uev <- unique(myData$EVTYPE)
flfl <- uev[grepl("flash", uev, ignore.case = TRUE)]
## Clean as following
myData[grepl("ICE STORM/FLASH FLOOD", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "ICE STORM"
myData[grepl("FLASH FLOOD WINDS", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl("FLASH FLOOD/", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl("FLASH FLOODS", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl("URBAN AND FLASH FLOODIN", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl("MUD SLIDES FLASH FLOOD", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl("FLASH FLOOD FROM ICE JAMS", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl("FLASH FLOOD - HEAVY RAIN", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl("FLOOD FLASH", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl("FLOOD/FLASH", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl("FLASH FLOOD LANDSLIDES", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl("FLASH FLOOD \\(MINOR", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl(" FLASH FLOOD", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl("URBAN AND FLASH FLOODIN", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
## There are observations with value FLOOD/FLASH FLOOD.  For purpose of this
## analysis that is being treated as FLOOD.
```
Work on events related to FLOOD.
```{r chkclnfld}
fld <- myData[grepl("^flood", myData$EVTYPE, ignore.case = TRUE), ]
## There are observations with value FLOODING.  These need to be 
## replaced with FLOOD.  Also, from above we know there are observations
## with values FLOOD/FLASH FLOOD.  These also need to be changed to FLOOD.
myData$EVTYPE <- sub("FLOODING", "FLOOD", myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("FLOOD/FLASH FLOOD", "FLOOD", 
                     myData$EVTYPE, ignore.case = TRUE)
fldp <- myData[grepl("flood", myData$EVTYPE, ignore.case = TRUE) &
                       !grepl("^flood", myData$EVTYPE, ignore.case = TRUE) &
                       !grepl("flash flood", myData$EVTYPE, ignore.case = TRUE) &
                       !grepl("coastal flood", myData$EVTYPE, 
                              ignore.case = TRUE) &
                       !grepl("lakeshore flood", myData$EVTYPE, 
                              ignore.case = TRUE), ]
## There are observations with values URBAN FLOOD and URBAN/SMALL STREAM FLOOD. 
## These are being classified as FLASH FLOOD for this analysis.  Also, there
## are observations with values RIVER FLOOD and BREAKUP FLOOD.  These need to
## be replaced by FLOOD.  Further replacements as indicated below:
## MINOR FLLOD -> FLOOD, MAJOR FLOOD -> FLOOD, RURAL FLOOD -> FLOOD
## ICE JAM FLOOD -> FLASH FLOOD, SMALL STREAM FLOOD -> FLASH FLOOD
## LAKE FLOOD -> LAKESHORE FLOOD, URBAN AND SMALL STREAM FLOODIN -> FLASH FLOOD
## RIVER AND STREAM FLOOD -> FLASH FLOOD, 
## HEAVY SNOW/HIGH WINDS & FLOOD -> HEAVY FLOOD,
## HEAVY RAIN AND FLOOD -> HEAVY RAIN,
## THUNDERSTORM WINDS/FLOOD -> THUNDERSTORM WIND, 
## Erosion/Cstl Flood -> COASTAL FLOOD, COASTAL  FLOOD/EROSION -> COASTAL FLOOD,
## TIDAL FLOOD -> FLOOD, THUNDERSTORM WINDS/ FLOOD -> THUNDERSTORM WIND
## 
myData$EVTYPE <- sub("RIVER FLOOD", "FLOOD", myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("BREAKUP FLOOD", "FLOOD", myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("URBAN FLOOD", "FLASH FLOOD", myData$EVTYPE, 
                     ignore.case = TRUE)
myData$EVTYPE <- sub("URBAN/SMALL STREAM FLOOD", "FLASH FLOOD", myData$EVTYPE, 
                     ignore.case = TRUE)
myData$EVTYPE <- sub("MAJOR FLOOD", "FLOOD", myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("MINOR FLOOD", "FLOOD", myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("RURAL FLOOD", "FLOOD", myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("ICE JAM FLOOD", "FLASH FLOOD", myData$EVTYPE, 
                     ignore.case = TRUE)
myData$EVTYPE <- sub("SMALL STREAM FLOOD", "FLASH FLOOD", myData$EVTYPE, 
                     ignore.case = TRUE)
myData$EVTYPE <- sub("LAKE FLOOD", "LAKESHORE FLOOD", myData$EVTYPE, 
                     ignore.case = TRUE)
myData$EVTYPE <- sub("URBAN AND SMALL STREAM FLOODIN", "FLASH FLOOD", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("RIVER AND STREAM FLOOD", "FLASH FLOOD", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("SNOWMELT FLOOD", "FLASH FLOOD", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("HEAVY SNOW/HIGH WINDS & FLOOD", "HEAVY SNOW", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("HEAVY RAIN AND FLOOD", "HEAVY RAIN", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("HEAVY RAINS/FLOOD", "HEAVY RAIN", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("THUNDERSTORM WINDS/FLOOD", "THUNDERSTORM WIND", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("Erosion/Cstl Flood", "COASTAL FLOOD", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("COASTAL  FLOOD/EROSION", "COASTAL FLOOD", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("TIDAL FLOOD", "FLOOD", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("THUNDERSTORM WINDS/ FLOOD", "THUNDERSTORM WIND", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("FLASH FLOOD/THUNDERSTORM WI", "FLASH FLOOD", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("FLOODS", "FLOOD", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("FLOOD/FLOOD", "FLOOD", 
                     myData$EVTYPE, ignore.case = TRUE)
## Check for more values
fldps <- myData[grepl("flood", myData$EVTYPE, ignore.case = TRUE) &
                       !grepl("^flood", myData$EVTYPE, ignore.case = TRUE) &
                       !grepl("flash flood", myData$EVTYPE, ignore.case = TRUE) &
                       !grepl("coastal flood", myData$EVTYPE, 
                              ignore.case = TRUE) &
                       !grepl("lakeshore flood", myData$EVTYPE, 
                              ignore.case = TRUE), ]
```
Check and clean events related to THUDERSTORM WIND.
```{r chkclntstm}
ts <- myData[grepl("^thun", myData$EVTYPE, ignore.case = TRUE), ]
## Based on output make following replacement 
## THUNDERSTORM WINDS -> THUNDERSTORM WIND, 
## THUNDERSTORM WINS -> THUNDERSTORM WIND
## Explore further
myData[grepl("^thun", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "THUNDERSTORM WIND"
tschk <- myData[grepl("^thun", myData$EVTYPE, ignore.case = TRUE) &
                        grepl("thunderstorm winds", myData$EVTYPE, 
                              ignore.case = TRUE), ]
## Look for rows having pattern thun but not in the start
tsb <- myData[grepl("thun", myData$EVTYPE, ignore.case = TRUE) &
                     !grepl("^thun", myData$EVTYPE, ignore.case = TRUE), ]
## Take care of observations with values like SEVERE THUNDERSTORM etc.
myData[grepl("severe thun", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "THUNDERSTORM WIND"
tsbchk <- myData[grepl("thun", myData$EVTYPE, ignore.case = TRUE) &
                     !grepl("^thun", myData$EVTYPE, ignore.case = TRUE), ]
## Clean following observations:
## LIGHTNING THUNDERSTORM WINDS -> THUNDERSTORM WIND
## LIGHTNING AND THUNDERSTORM WIN -> THUNDERSTORM WIND
myData[grepl("LIGHTNING THUNDERSTORM WINDS", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "THUNDERSTORM WIND"
myData[grepl("LIGHTNING AND THUNDERSTORM WIN", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "THUNDERSTORM WIND"
myData[grepl("TUNDERSTORM WIND", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "THUNDERSTORM WIND"
myData[grepl("THUDERSTORM WINDS", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "THUNDERSTORM WIND"
## There are observations that have TSTM as short form of THUNDERSTORM.  Replace
## entries having TSTM with THUNDERSTORM WIND
## check to see more entries starting with TSTM
myData[grepl("^tstm", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "THUNDERSTORM WIND"
tstm <- myData[grepl("TSTM", myData$EVTYPE, ignore.case = TRUE), ]
## There are following observations that need to be cleaned
## NON-TSTM WIND -> HIGH WIND
## '  TSTM WIND' -> THUNDERSTORM WIND, NON-TSTM WIND -> HIGH WIND
## '  TSTM WIND (' -> THUNDERSTORM WIND
## MARINE TSTM WIND -> MARINE THUNDERSTORM WIND
myData[grepl("MARINE TSTM WIND", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "MARINE THUNDERSTORM WIND"
myData[grepl("NON-TSTM WIND", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "HIGH WIND"
myData[grepl("TSTM WIND", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "THUNDERSTORM WIND"
tstmchk <- myData[grepl("TSTM", myData$EVTYPE, ignore.case = TRUE), ]
```
Work on events related to FROST/FREEZE.
```{r chkclnfrst}
uev <- unique(myData$EVTYPE)
frst <- uev[grepl("frost", uev, ignore.case = TRUE)]
## Five values of EVTYPE containing FROST.  Change all of them to FROST/FREEZE.
myData[grepl("frost", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FROST/FREEZE"
myData[grepl("freeze", myData$EVTYPE, 
                    ignore.case = TRUE) &
               !grepl("FROST/FREEZE", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FROST/FREEZE"
```
Work on events related to HAIL.
```{r chkclnhl}
uev <- unique(myData$EVTYPE)
hl <- uev[grepl("hail", uev, ignore.case = TRUE)]
## 19 values of EVTYPE containing HAIL.  All except MARINE HAIL need to be
## replaced with HAIL.
myData[grepl("^hail", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "HAIL"
myData[grepl("WIND/HAIL", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "HAIL"
myData[grepl("SMALL HAIL", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "HAIL"
myData[grepl("GUSTY WIND/HAIL", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "HAIL"
```
Work on events related to HEAT.
```{r chkclnht}
uev <- unique(myData$EVTYPE)
ht <- uev[grepl("heat", uev, ignore.case = TRUE)]
## 7 values of EVTYPE containing heat.  Change all of them to HEAT except for
## excessive heat.
myData[grepl("heat", myData$EVTYPE, 
                    ignore.case = TRUE) &
               !grepl("excessive heat", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "HEAT"
```
Work on events related to HEAVY RAIN.
```{r chkclnhvrn}
uev <- unique(myData$EVTYPE)
hvrn <- uev[grepl("rain", uev, ignore.case = TRUE)]
## 34 values of EVTYPE containing rain.  For this analysis change all of 
## them to HEAVY RAIN.
myData[grepl("rain", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "HEAVY RAIN"
```
Work on events related to HEAVY SNOW.
```{r chkclnhvsn}
uev <- unique(myData$EVTYPE)
hvsn <- uev[grepl("snow", uev, ignore.case = TRUE)]
## 49 values of EVTYPE containing rain.  For this analysis change all of 
## them to HEAVY SNOW except for thos pertaining to LAKE-EFFECT SNOW.
myData[grepl("snow", myData$EVTYPE, 
                    ignore.case = TRUE) &
               !grepl("lake", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "HEAVY SNOW"
```
Work on events related to LAKE-EFFECT SNOW.
```{r chkclnlksn}
uev <- unique(myData$EVTYPE)
lksn <- uev[grepl("lake", uev, ignore.case = TRUE)]
## HEAVY LAKE SNOW -> LAKE-EFFECT SNOW,  Lake Effect Snow -> LAKE-EFFECT SNOW 
myData[grepl("HEAVY LAKE SNOW", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "LAKE-EFFECT SNOW"
myData[grepl("Lake Effect Snow", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "LAKE-EFFECT SNOW"
```
Work on events related to HIGH SURF.
```{r chkclnhsrf}
uev <- unique(myData$EVTYPE)
hsrf <- uev[grepl("tide", uev, ignore.case = TRUE)]
## 11 values of EVTYPE containing surf.  For this analysis change all of
## them to HIGH SURF
myData[grepl("surf", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "HIGH SURF"
```
Work on events related to HIGH WIND.
```{r chkclnhwnd}
uev <- unique(myData$EVTYPE)
wnd <- uev[grepl("wind", uev, ignore.case = TRUE)]
## 35 values of EVTYPE containing wind
## Disregard Cold/Wind Chill, Extreme Cold/Wind Chill, High Wind, Marine,
## Strong wind, Thunderstorm Wind, gusty wind, HURRICANE OPAL and WINTER STORM.
## Replace remaining with HIGH WIND.
uev <- unique(myData$EVTYPE)
clnwnd <- uev[grepl("wind", uev, ignore.case = TRUE) &
                   !grepl("HURRICANE OPAL", uev, ignore.case = TRUE) &
                   !grepl("MARINE", uev, ignore.case = TRUE) &
                   !grepl("WINTER STORM", uev, ignore.case = TRUE) &
                   !grepl("Cold/Wind Chill", uev, ignore.case = TRUE) &
                   !grepl("Extreme Cold/Wind Chill", uev, ignore.case = TRUE) &
                   !grepl("Strong wind", uev, ignore.case = TRUE) &
                   !grepl("Thunderstorm Wind", uev, ignore.case = TRUE) &
                   !grepl("gusty wind", uev, ignore.case = TRUE)]
myData[grepl("wind", myData$EVTYPE, 
                    ignore.case = TRUE) &
               !grepl("HURRICANE OPAL", myData$EVTYPE, 
                    ignore.case = TRUE) &
               !grepl("MARINE", myData$EVTYPE, 
                    ignore.case = TRUE) &
               !grepl("WINTER STORM", myData$EVTYPE, 
                    ignore.case = TRUE) &
               !grepl("Cold/Wind Chill", myData$EVTYPE, 
                      ignore.case = TRUE) &
               !grepl("Extreme Cold/Wind Chill", myData$EVTYPE, 
                      ignore.case = TRUE) &
               !grepl("Strong wind", myData$EVTYPE, 
                      ignore.case = TRUE) &
               !grepl("Thunderstorm Wind", myData$EVTYPE, 
                      ignore.case = TRUE) &
               !grepl("gusty wind", myData$EVTYPE, 
                      ignore.case = TRUE), "EVTYPE"] <- "HIGH WIND"

```
Work on events related to HURRICANE (TYPHOON)
```{r chkclnhrcn}
uev <- unique(myData$EVTYPE)
hrcn <- uev[grepl("hur", uev, ignore.case = TRUE)]
## 10 values of EVTYPE containing hur.  For this analysis change all of
## them to HURRICANE (TYPHOON)
myData[grepl("hur", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "HURRICANE (TYPHOON)"
myData[grepl("TYPHOON", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "HURRICANE (TYPHOON)"
```
Work on events related to ICE STORM.
```{r chkclnicst}
uev <- unique(myData$EVTYPE)
icst <- uev[grepl("ice", uev, ignore.case = TRUE)]
## 11 values of EVTYPE containing ice.  For this analysis change all of
## them to ICE STORM as they seem pertaining to an ice storm.
myData[grepl("ice", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "ICE STORM"
```
Work on events related to LIGHTNING.
```{r chkclnltng}
uev <- unique(myData$EVTYPE)
ltng <- uev[grepl("light", uev, ignore.case = TRUE)]
## 6 values of EVTYPE containing ice.  Change all of
## them to LIGHTNING.
myData[grepl("light", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "LIGHTNING"
myData[grepl("LIGNTNING", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "LIGHTNING"
```
Work on events related to MARINE HIGH WIND.
```{r chkclnmrn}
uev <- unique(myData$EVTYPE)
mrn <- uev[grepl("marine", uev, ignore.case = TRUE)]
## 2 values of EVTYPE containing don't comply.  Change both of
## them to MARINE HIGH WIND for the purpose of this analysis.
myData[grepl("Marine Accident", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "MARINE HIGH WIND"
myData[grepl("MARINE MISHAP", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "MARINE HIGH WIND"
```
Work on events related to RIP CURRENT.
```{r chkclnrpcr}
uev <- unique(myData$EVTYPE)
rpcr <- uev[grepl("rip", uev, ignore.case = TRUE)]
## Values RIP CURRENTS of EVTYPE to be changed to RIP CURRENT.  
myData[grepl("RIP CURRENTS", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "RIP CURRENT"
```
Work on events related to STORM SURGE/TIDE.
```{r chkclnstsg}
uev <- unique(myData$EVTYPE)
stsg <- uev[grepl("Surge", uev, ignore.case = TRUE)]
## Two Values STORM SURGE and "COASTAL SURGE of EVTYPE need to be made compliant  
myData[grepl("STORM SURGE", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "STORM SURGE/TIDE"
myData[grepl("COASTAL SURGE", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "STORM SURGE/TIDE"
```
Work on events related to STRONG WIND.
```{r chkclnstwnd}
uev <- unique(myData$EVTYPE)
stwnd <- uev[grepl("Strong", uev, ignore.case = TRUE)]
## Values Strong winds of EVTYPE need to be made compliant  
myData[grepl("Strong winds", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "STRONG WIND"
uev <- unique(myData$EVTYPE)
gswnd <- uev[grepl("gusty", uev, ignore.case = TRUE)]
## Values gusty winds of EVTYPE need to be made compliant  
myData[grepl("gusty", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "STRONG WIND"
```
Work on events related to TROPICAL STORM.
```{r chkclntrst}
uev <- unique(myData$EVTYPE)
trst <- uev[grepl("trop", uev, ignore.case = TRUE)]
## All values having "tRop" except TROPICAL DEPRESSION of EVTYPE need to be made
## compliant  
myData[grepl("trop", myData$EVTYPE, ignore.case = TRUE) &
               !grepl("tropical depression", myData$EVTYPE, 
                      ignore.case = TRUE), "EVTYPE"] <- "TROPICAL STORM"
```
Work on events related to WATERSPOUT.
```{r chkclnws}
uev <- unique(myData$EVTYPE)
ws <- uev[grepl("water", uev, ignore.case = TRUE)]
## All values having "water" need to be made
## compliant  
myData[grepl("water", myData$EVTYPE, 
             ignore.case = TRUE), "EVTYPE"] <- "WATERSPOUT"
```
Work on events related to WILDFIRE.
```{r chkclnwf}
uev <- unique(myData$EVTYPE)
wf <- uev[grepl("wild", uev, ignore.case = TRUE)]
## All values having "wild" need to be made
## compliant  
myData[grepl("wild", myData$EVTYPE, 
             ignore.case = TRUE), "EVTYPE"] <- "WILDFIRE"
```
Work on events related to WINTER WEATHER.
```{r chkclnwth}
uev <- unique(myData$EVTYPE)
wth <- uev[grepl("weather", uev, ignore.case = TRUE)]
## All values having "weather" except "warm weather" need to be made
## compliant.  Warm weather will be changed to heat.  
myData[grepl("WARM WEATHER", myData$EVTYPE, 
             ignore.case = TRUE), "EVTYPE"] <- "HEAT"
myData[grepl("WEATHER", myData$EVTYPE, 
             ignore.case = TRUE), "EVTYPE"] <- "WINTER WEATHER"
```
Work on events related to WINTER STORM.
```{r chkclnwst}
uev <- unique(myData$EVTYPE)
wtst <- uev[grepl("winter", uev, ignore.case = TRUE)]
## All values having "winter" except "winter weather" need to be made
## compliant.  
myData[grepl("winter", myData$EVTYPE, 
             ignore.case = TRUE) &
               !grepl("winter weather", myData$EVTYPE, 
                      ignore.case = TRUE), "EVTYPE"] <- "WINTER STORM"
```
Rest of the values were taken care of while working on events specified above.

Convert all events to be in upper case for consistency.
```{r ucase}
myData$EVTYPE <- toupper(myData$EVTYPE)
```

## Results

### Find which weather events are most harmful with respect to population health across the United States

The idea is to show the impact of weather events first in terms of Injuries and fatalities individually and then the combined impact of both injuries and fatalaties.  In order to show impact in terms of injuries and fatalaties individually, a panel plot is used.  The combined impact is depicted by a plot of its own.
``` {r health}
## Plot injuries and fatalaties
library(ggplot2)
## Analyzing injuries
inj <- myData[,c("EVTYPE", "INJURIES")]
## Total injuries for each event type
eventinj <- aggregate(INJURIES ~ EVTYPE, data=inj, FUN=sum)
## Order the injuries in descending order.
eventinj <- eventinj[order(eventinj$INJURIES, decreasing = TRUE),]
## Take 15 events that caused most injuries for this analysis
eventinj <- eventinj[1:15, ]
## Analyzing fatalities
ftlts <- myData[,c("EVTYPE", "FATALITIES")]
## Total fatalities for each event type
eventftl <- aggregate(FATALITIES ~ EVTYPE, data=ftlts, FUN=sum)
eventftl <- eventftl[order(eventftl$FATALITIES, decreasing = TRUE),]
## Take 15 events that caused most FATALITIES for this analysis
eventftl <- eventftl[1:15, ]
##Panel plot
injvec <- vector(mode="character", length=15)
injvec[1:15] <- "Injuries"
treventinj <- data.frame("Events"=eventinj$EVTYPE, 
                         "Number" = eventinj$INJURIES, 
                         "Consequence" = injvec)
ftlvec <- vector(mode="character", length=15)
ftlvec[1:15] <- "Fatalities"
treventftl <- data.frame("Events"=eventftl$EVTYPE, 
                         "Number" = eventftl$FATALITIES, 
                         "Consequence" = ftlvec)
combevent <- rbind(treventinj, treventftl)
combevent
ggplot(data=combevent, aes(x=Events, y=Number, fill=Events)) + 
        geom_bar(stat="identity") +
        xlab("Weather event") + ylab("Injuries/Fatalities per weather event") +
        ggtitle("Injuries and Fatalities by weather event in US") +
        theme(axis.text.x  = element_text(angle=90, vjust=0.5, hjust=1)) +
        facet_grid(Consequence ~ .)
## Take into consideration sum of injuries and fatalities
suminft <- rowSums(cbind(myData$INJURIES, myData$FATALITIES))
tothlth <- data.frame("EVTYPE" = myData$EVTYPE, 
                 "HEALTH" = suminft)
## Sum of fatalities and injuries for each event type
eventsm <- aggregate(HEALTH ~ EVTYPE, data=tothlth, FUN=sum)
##Order descending
eventsm <- eventsm[order(eventsm$HEALTH, decreasing = TRUE),]
## Take 15 events that caused most FATALITIES for this analysis
eventsm <- eventsm[1:15, ]
eventsm
##Plot sum of injuries and fatalities per event
ggplot(data=eventsm, aes(x=EVTYPE, y=HEALTH, fill=EVTYPE)) + 
        geom_bar(stat="identity") +
        xlab("Weather event") + 
        ylab("Sum of Injuries and Fatalities per event") +
        ggtitle("Sum of Injuries and Fatalities by weather event across US") +
        theme(axis.text.x  = element_text(angle=90, vjust=0.5, hjust=1))
```

From this analysis we can see that **TORNADO** was the event that caused most harm in terms of both injuries and fatalities individually as well as injuries and fatalities combined.

### Assess across the United Statess, which weather event caused most damage to economy implying the most damage caused in terms of both Property Damage as well as Crop Damage

The strategy is to get the data in form where we can make the above assessment and depict it on a plot to get a sense of the assessment.  Most of the transformations involve calculating total damage so that we can make the assessment.
```{r expleconomy}
## Explore data
eco <- myData[, c("EVTYPE", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
unqc <- unique(myData$CROPDMGEXP)
unqp <- unique(myData$PROPDMGEXP)
```
Looking at the data, columns CROPDMGEXP and PROPDMGEXP contain values 0-9 and one of B,b,M,m,K,k,H,h.  There are other values like -,?,+.  These values will be ignored because they are not relevant for exponent field.  We need use appropriate conversion strategy for rest of them.  Since the highest is B,b (Billion), taking that as the unit, implies we consider billion to be the 0 of the scale.  So, 0 becomes 1e-09 to represent Billion.  Similarly, million becomes 1e-03 in terms of billions and so on.  This is the conversion strategy used here.  Once we convert the exponent fields and add columns for corresponding converted values, we need to multiply the conversion fields to their respective DMG field (multiply CROPDMG with CROPDMGEXPCONV and PROPDMG with PROPDMGEXPCONV).
```{r convexp}
#base vector
base <- c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, "b", "B", "m", "M", "k", "K", "h", "H")
#conversion vector
conv <- c(1e-09, 1e-08, 1e-07, 1e-06, 1e-05, 1e-04, 1e-03, 0.01, 0.1, 1, 1, 1,
        1e-03, 1e-03, 1e-06, 1e-06, 1e-07, 1e-07)
##Create a data frame representing a conversion table
convtab <- data.frame(base=base, conv=conv)
## Subset the dataframe to contain values from bs only in PROPDMGEXP and
## CROPDMGEXP
clneco <- subset(eco, eco$CROPDMGEXP %in% convtab$base | 
                         eco$PROPDMGEXP %in% convtab$base)
## Add column for conversion of CROPDMNEXP
clneco$CROPDMGEXPCONV <- sapply(clneco$CROPDMGEXP, function(x)
        if(x %in% convtab$base) convtab[base == x, "conv"] else 0)
## Add column for conversion of PROPDMNEXP
clneco$PROPDMGEXPCONV <- sapply(clneco$PROPDMGEXP, function(x)
        if(x %in% convtab$base) convtab[base == x, 2] else 0)
```
Now multipy the new columns CROPDMG and PROPDMG with CROPDMGEXPCONV and PROPDMGEXPCONV respectively to get corresponding damage in billions.  Then add these damage values in billions for CROPDMG and PROPDMG to get total damage.
```{r caldmg}
clneco <- transform(clneco, TOTDMG = CROPDMG * CROPDMGEXPCONV + 
                  PROPDMG * PROPDMGEXPCONV)
```
Calculate total damage per weather event and then get 15 events that caused maximum damage in the decreasing order.  Here the inflation rate is not taken into account so the deflation in the value of money is ignored.
```{r eventdmg}
## Subset dataframe to contain event and total damage only
dmg <- data.frame("Event" = clneco$EVTYPE, "Damage" = as.numeric(clneco$TOTDMG))
eventdmg <- aggregate(Damage ~ Event, data=dmg, FUN=sum)
eventdmg <- eventdmg[order(eventdmg$Damage, decreasing = TRUE),]
## Take 15 events that caused most Damage for this analysis
eventdmg <- eventdmg[1:15, ]
eventdmg
```
Create a bar plot to get sense of 15 most damaging weather events.
```{r dmgplot}
library(ggplot2)
ggplot(data=eventdmg, aes(x=Event, y=Damage, fill=Event)) + 
        geom_bar(stat="identity") +
        xlab("Weather event") + 
        ylab("Total damage in Billions per weather event") +
        ggtitle("Total damage in billions by weather event across US") +
        theme(axis.text.x  = element_text(angle=90, vjust=0.5, hjust=1))
```

From this analysis we can see that **FLOOD** caused most damage in economic terms.